\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-05-22 forwards-and-backwards.tex
angle bracket process; more citations

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which
the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.
Equations of this form have been studied extensively
in the context of spatial ecology;
see for instance \citet{lam2023introduction} and \citet{cantrell2004spatial},
and in many other fields;
for instance, \citet{ghosh2022emergent}
derive a stochastic version of~\eqref{semilinear PDE} to describe abundances of mutant bacteria strains
along the human gut,
while \citet{li/buenzli/simpson:2022} study the effects of nonlinear diffusion
on long-term survival of a lattice-based interacting particle system.
% MORE DETAILS:
% \cite{li/buenzli/simpson:2022} write down an interacting particle
% system on a hexagonal lattice (with at most one individual per site) that incorporates an effect
% of local crowding on individual motion. Their equation
% corresponds to a reaction-diffusion equation with nonlinear diffusion and a 
% bistable reaction term of the form~(\ref{Birzu nonlinearity}) and they
% investigate the interplay between nonlinear diffusion and long-term 
% survival or extinction of the population.

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:mgale_construction}
	For each value of $N$ and $\theta$, and each purely atomic 
$\eta_0^N\in\measures$ with atoms of mass $1/N$,
$(\eta^N_t)_{t \geq 0}$ is the (scaled) empirical measure of a birth-death 
process with c\`adl\`ag paths in $\measures$ for which,
for all $f \in C^{\infty}_{b}(\IR^d)$,
writing $q_\theta(x, dy)$ for the Gaussian kernel
with mean $x + \meanq(x)/\theta$ and covariance $\covq(x) / \theta$,
\begin{equation}
    \label{eqn:eta_martingale}
\begin{aligned}
M^N_t(f)
&:=  \langle f, \eta^N_t \rangle
        -\langle f, \eta^N_0 \rangle
 \\ &\qquad {}
 -  \int_{0}^{t}\bigg\{
        \bigg\langle \left( \int \theta
             \left(
                f(z)r(z, \eta^N_{s})
                - f(x)r(x, \eta^N_{s})
            \right)
        q_\theta(x,dz) \right)
%\\ & \qquad \qquad \qquad \qquad {}
         \gamma(x, \eta^N_{s}), \eta^N_{s}(dx) \bigg\rangle
\\ & \qquad \qquad \qquad \qquad \qquad {}
    + \bigg\langle f(x) F(x, \eta^N_{s}),
    \eta^N_{s}(dx) \bigg\rangle
    \bigg\} ds
\end{aligned}    
\end{equation}
is a martingale (with respect to the natural filtration), 
with quadratic variation
    \begin{equation} \label{eqn:prelimit_martingale_variation}
\begin{aligned} \relax
[ M^N(f)]_t =& 
\frac{\theta}{N} \int_{0}^{t}\bigg\{
    \Big\langle \gamma(x, \eta^N_{s})
        \int f^2(z)r(z, \eta^N_{s}) q_\theta(x,dz) 
    , \eta^N_{s}(dx) \Big\rangle \\
& \qquad \qquad {}
    + \Big\langle \mu_\theta(x, \eta^N_{s}) f^2(x) 
    , \eta^N_{s}(dx)\Big\rangle
    \bigg\}ds. 
\end{aligned}    
\end{equation}
\end{definition}
The form of~(\ref{eqn:eta_martingale}) and~(\ref{eqn:prelimit_martingale_variation})
is explained in Section~\ref{sec:heuristics}.
Note that since (juvenile) individuals are produced at rate $N \gamma \eta$, but each
has mass $1/N$,
these factors of $N$ cancel in~\eqref{eqn:eta_martingale}. 
Under our scaling, $N$ and $\theta=\theta(N)$ will tend to infinity in such a 
way that $\alpha:=\lim_{N\to\infty}\theta(N)/N$ exists and is finite.
From the expression~(\ref{eqn:prelimit_martingale_variation}) it is easy to guess that 
whether the limiting processes will be deterministic or stochastic
is determined by whether $\alpha$ is zero or nonzero.

@@

Since we take bounded $f$,
for most situations the bound $C_f(1+|f(x)|)$ in condition~\ref{gamma_B_condition} above
can be safely replaced simply by $C_f$;
however, this will be useful
in certain situations where we consider a sequence of $f$ with increasing upper bounds.
We now give two concrete situations in which condition~\ref{gamma_B_condition} is satisfied.
The proof is in Section~\ref{sec:preliminary_proofs}.

\begin{lemma}
    \label{lem:conditions on r}
    Assume that Conditions~\ref{def:model_setup} are satisfied,
    except for condition~\ref{gamma_B_condition}.
    If either
\begin{enumerate}
\item
\label{control through r} 
    $|\partial_{x_i}r(x,\eta)|$ and 
    $|\partial_{x_ix_j}r(x,\eta)|$
    are uniformly bounded for $x\in\IR^d$, $\eta\in\measures$;
\item 
\label{control through gamma}
    or, $m^2\gamma(x,m)$ is uniformly bounded and there exists $C<\infty$
    such that for $\theta$ sufficiently large, and all $x\in\IR^d$, $\eta\in\measures$,
    \[
    \theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)q_\theta(x,dy)\leq C\smooth{\gamma}\eta(x),
    \]
    and
    \[\theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)^2q_\theta(x,dy)
    \leq C(\smooth{\gamma}\eta(x))^2 ,
    \]
\end{enumerate} 
    then condition~\ref{gamma_B_condition} is also satisfied.
\end{lemma}

@@

\begin{theorem} \label{thm:nonlocal_convergence}
    Let $(\eta^N_t)_{t \geq 0}$
    be as defined in Definition~\ref{defn:mgale_construction}
	and assume that as $N \to \infty$, $\theta(N) \to \infty$
	in such a way that $\theta(N)/N \to \alpha$.
    (However, the kernels $\rho_r$, $\rho_\gamma$, and $\rho_F$
    remain fixed.) 
    Suppose that Assumptions~\ref{def:model_setup} hold and, 
    further, that $\{\eta_0^N\}_{N\geq 1}$ is a sequence of 
    purely atomic measures, with $\eta_0^N$ comprised of atoms of mass $1/N$,
    which is tight in $\measures$.
    Also assume there exists a nonnegative $f_0 \in C(\IR^d)$
    with uniformly bounded first and second derivatives
    (i.e., with $\sup_x \sup_{\|z\|=1} \sum_i \partial_{x_i} f_0(x) z_i$
    and $\sup_x \sup_{\|z\|=1} \sum_{ij} \partial_{x_i x_j} f_0(x) z_i z_j$ both finite)
    and $f_0(x) \to \infty$ as $|x| \to \infty$ for which
    $\langle f_0(x), \eta_0^N(dx)\rangle<C<\infty$
    for some $C$ independent of $N$.
    Then the sequence of processes
    $(\eta^N_t)_{t \geq 0}$ is tight, and for any limit point
    $(\eta_t)_{t \geq 0}$, for every $f \in C^\infty_b(\IR^d)$,
    \begin{align} \label{eqn:limiting_mgale_problem}
        \begin{split}
        M_t(f)
            &:=
            \langle f(x), \eta_t(dx) \rangle
            -
            \langle f(x), \eta_0(dx) \rangle
            \\ & \qquad
            -
            \int_0^t \big\langle
                \gamma(x, \eta_s)
                \mathcal{B}\left(
                    f(\cdot) r(\cdot, \eta_s)
                \right)(x)
                %\\ &\qquad \qquad \qquad {}
                +
                f(x)
                F(x, \eta_s),
                \eta_{s}(dx)
            \big\rangle ds
        \end{split}
    \end{align}
    is a martingale (with respect to the natural filtration),
    with quadratic variation
    \begin{align} \label{eqn:limiting_mgale_variation}
        [ M(f) ]_t
        =
        \alpha
        \int_0^t
        \big\langle
            %\left(
	    2\gamma\left( x, \eta_{s} \right)
            r\left(x, \eta_{s} \right)% + \mu\left(x, \eta_{s} \right)\right)
            f^2(x),
            \eta_{s} (dx)
        \big\rangle ds. 
    \end{align}
    If $\alpha = 0$ the limit is deterministic.
\end{theorem}

@@

We now look at the second order term in~(\ref{generator prelimit}).
An individual at location $x$ gives birth 
to a surviving offspring at $y$ at rate
$$
\gamma(x,\eta) r(y,\eta) q_{\theta}(x, dy) ,
$$
and since this increments $\langle f, \eta \rangle$ by $f(y) / N$,
the contribution to the quadratic variation from birth events,
which occur at rate $\theta$ per individual 
(so, rate $N\theta |\eta|$ overall), is
$$
\int
    N \theta \gamma(x,\eta)
    \int \frac{1}{N^2} f^2(y) r(y,\eta)
    q_\theta(x,dy) 
\eta(dx) .
$$
Similarly, the increment in $\langle f, \eta\rangle$ resulting from 
the death of an individual at $x$ is $-f(x)/N$, and so combining with the 
above, the second order term in the generator takes the form
\begin{align*}
& G''(\langle f,\eta\rangle)
\frac{1}{2} N \theta
\left\{
    \int
        \gamma(x,\eta)
        \int \frac{1}{N^2}f^2(y)r(y,\eta)q_\theta(x,dy) 
    \eta(dx)
    +
    \int
        \mu_\theta(x,\eta)\frac{1}{N^2}f^2(x) 
    \eta(dx)
\right\} \\
&\qquad
= \frac{1}{2} G''(\langle f, \eta \rangle)
    \frac{\theta}{N}
	\int\left\{
		\gamma(x, \eta) \int f^2(y) r(y, \eta) q_\theta(x, dy) + f^2(x) \mu_\theta(x, \eta)
    \right\} \eta(dx) .
\end{align*}
Since $\int f^2(y) r(y, \eta) q_\theta(x, dy) \to f^2(x) r(x, \eta)$
and $r\gamma  + \mu_\theta = 2 r \gamma  - F / \theta \to 2 r \gamma$
as $\theta \to \infty$,
this converges to
\begin{align*}
\frac{\alpha}{2} G''(\langle f, \eta \rangle)
    \big\langle
	2 r(x, \eta) \gamma(x, \eta),
        \eta(dx)
    \big\rangle .
\end{align*}

@@

\begin{proof}
    First note that by Lemma~\ref{lem:eta_f_bound},
    $\IE[\langle 1, \eta^N_t \rangle] \le \IE[\langle 1, \eta^N_0 \rangle] e^{Ct}$
    for some $C$ (independent of $N$).
    Now, let $M^{N*}_t(f) = \sup_{0 \le s \le t} M^N_t(f)$,
    and as before let $[M^N(f)]_t$ be the quadratic variation of $M^N_t(f)$,
    so that by the Burkholder-Davis-Gundy inequality, there is a $C'$ so that,
    in the notation of Lemma~\ref{lem:eta_f_bound},
    \begin{align*}
        \IE\left[ M^{N*}_t(1) \right]
        &\le
        C' \IE\left[ \sqrt{[M^N(1)]_t} \right]
        \\ &\le
        C'\left( 1 + \IE\left[ [M^N(1)]_t \right] \right)
        \\ &=
        C'\left( 1 + \frac{\theta}{N} \IE\left[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\left\{
                \gamma(x, \eta^N_s)
                \int_{\IR^d} r(y, \eta^N_s) q_\theta(x, dy)
                + \mu_\theta(x, \eta^N_s)
            \right\}, \eta_s^N(dx)\Big\rangle
            ds
            \right] \right)
        \\ &=
        C'\Big( 1 + \IE\Big[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\Big\{
                \frac{2\theta}{N} \gamma(x, \eta^N_s) r(x, \eta^N_s)
        \\ &\qquad\qquad\qquad\qquad \qquad {}
		+ \frac{\gamma(x,\eta_s^N)}{N} 
		B^\theta_1(x, \eta^N_s)-\frac{1}{N}F(x,\eta^N_s)
            \Big\}, \eta_s^N(dx) \Big\rangle
            ds
            \Big] \Big) .
    \end{align*}

@@

To see why $\psi_t^{\epsilon,x}$ is useful,
first note that for any time-dependent function $\phi_t(x)$
with time derivative $\dot \phi_t(x) = \partial_t \phi_t(x)$,
\begin{multline}
\label{time dep mg prob}
    \langle \phi_t(x), \eta_t(dx) \rangle
    = 
    \langle \phi_0(x), \eta_0(dx) \rangle
    +
    M_t(\phi)
    +
    \int_0^t \big\langle \mathcal{L}^\theta \phi_s(x) 
        + \dot \phi_s(x), \eta_s(dx) \big\rangle ds
    \\  {}
    +
    \int_{0}^t \big\langle \phi_s(x) F(x, \eta_s) , \eta_s(dx) \big\rangle ds ,
\end{multline}
where $M_t(\phi)$ is a martingale (with respect to the natural filtration)
with quadratic variation given by~\eqref{eqn:prelimit_martingale_variation}
with $f$ replaced by $\phi_s(\cdot)$.
So, taking $\phi_s(\cdot)=\psi_{t-s}^{\epsilon,x}(\cdot)$ for $0 \le s \le t$,
\begin{eqnarray}
    \nonumber
    \rho_\epsilon*\eta_t(x)&=&\langle\psi^{\epsilon,x}_0(y),\eta_t(dy)\rangle
    \\
    &=&\langle\psi^{\epsilon,x}_t(y),\eta_0(dy)\rangle
    +\int_0^t\big\langle\psi^{\epsilon,x}_{t-s}(y)
    F\big(\rho_\epsilon*\eta_s(y)\big), \eta_s(dy)\big\rangle ds
    +M_t(x),
    \label{expn rhoepsilon} 
\end{eqnarray}
where $M_t(x)$ has mean zero and a second moment we can easily write down.

@@

Consulting~\eqref{eqn:prelimit_martingale_variation},
and using the approach of Lemma~\ref{lem:limit_mgale},
the quadratic variation of $\widehat{M}^N(f)$ satisfies
$ \IE[[\widehat{M}^N_f]_T] \le C (\theta/N) \int_0^T \IE[\langle 1, \eta_s \rangle] ds \le C' \theta/N$
for some constants $C$ and $C'$.
Now, by the Burkholder-Davis-Gundy inequality,
$\IE[\sup_{0 \le u \le T} |\widehat{M}^N(f)_u|^2] \le C'' \IE[[\widehat{M}^N(f)]_u]$,
and so using Markov's inequality,
\begin{equation}
\label{martingale term to zero}
    \limsup_{N\to\infty}\IP\Big[
        2\sup_{0\leq u\leq T}|\widehat{M}^N(f)_u|
        >\frac{\nu}{3}
    \Big]
    \leq
    \limsup_{N\to\infty} \frac{36}{\nu^2} C'' \IE\big[[\widehat{M}^N(f)]_T\big]
    \leq
    \limsup_{N\to\infty}\frac{36}{\nu^2}\frac{C' C''\theta}{N\epsilon^d} 
    =0. 
\end{equation}

>>>>

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which
the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.
Equations of this form have been studied extensively
in the context of spatial ecology
(see for instance \citet{lam2023introduction} and \citet{cantrell2004spatial})
and in many other fields;
for instance, \citet{ghosh2022emergent}
derive a stochastic version of~\eqref{semilinear PDE} to describe abundances of mutant bacteria strains
along the human gut,
while \citet{li/buenzli/simpson:2022} study the effects of nonlinear diffusion
on long-term survival of a lattice-based interacting particle system.
However, we do not study the effect of movement of adults,
which can additionally affect the limiting equations:
see for instance \citet{holmes/lewis/banks/veit:1994} or \citet{potts/borger:2023}.
% MORE DETAILS:
% \cite{li/buenzli/simpson:2022} write down an interacting particle
% system on a hexagonal lattice (with at most one individual per site) that incorporates an effect
% of local crowding on individual motion. Their equation
% corresponds to a reaction-diffusion equation with nonlinear diffusion and a 
% bistable reaction term of the form~(\ref{Birzu nonlinearity}) and they
% investigate the interplay between nonlinear diffusion and long-term 
% survival or extinction of the population.

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:mgale_construction}
	For each value of $N$ and $\theta$, and each purely atomic 
$\eta_0^N\in\measures$ with atoms of mass $1/N$,
$(\eta^N_t)_{t \geq 0}$ is the (scaled) empirical measure of a birth-death 
process with c\`adl\`ag paths in $\measures$ for which,
for all $f \in C^{\infty}_{b}(\IR^d)$,
writing $q_\theta(x, dy)$ for the Gaussian kernel
with mean $x + \meanq(x)/\theta$ and covariance $\covq(x) / \theta$,
\begin{equation}
    \label{eqn:eta_martingale}
\begin{aligned}
M^N_t(f)
&:=  \langle f, \eta^N_t \rangle
        -\langle f, \eta^N_0 \rangle
 \\ &\qquad {}
 -  \int_{0}^{t}\bigg\{
        \bigg\langle \left( \int \theta
             \left(
                f(z)r(z, \eta^N_{s})
                - f(x)r(x, \eta^N_{s})
            \right)
        q_\theta(x,dz) \right)
%\\ & \qquad \qquad \qquad \qquad {}
         \gamma(x, \eta^N_{s}), \eta^N_{s}(dx) \bigg\rangle
\\ & \qquad \qquad \qquad \qquad \qquad {}
    + \bigg\langle f(x) F(x, \eta^N_{s}),
    \eta^N_{s}(dx) \bigg\rangle
    \bigg\} ds
\end{aligned}    
\end{equation}
is a martingale (with respect to the natural filtration), 
with angle bracket process
    \begin{equation} \label{eqn:prelimit_martingale_variation}
\begin{aligned} \relax
\left\langle M^N(f) \right\rangle_t =& 
\frac{\theta}{N} \int_{0}^{t}\bigg\{
    \Big\langle \gamma(x, \eta^N_{s})
        \int f^2(z)r(z, \eta^N_{s}) q_\theta(x,dz) 
    , \eta^N_{s}(dx) \Big\rangle \\
& \qquad \qquad {}
    + \Big\langle \mu_\theta(x, \eta^N_{s}) f^2(x) 
    , \eta^N_{s}(dx)\Big\rangle
    \bigg\}ds. 
\end{aligned}    
\end{equation}
\end{definition}

The angle bracket process (or, ``conditional quadratic variation'')
is the unique previsible process making $(M^N(f)_t)^2 - \langle M^N(f) \rangle_t$ a martingale
with respect to the natural filtration.
It differs from the usual quadratic variation
(usually denoted $[M^N(f)]_t$)
because the process has jumps;
for the (continuous) limit the two notions will coincide.
The use of angle brackets for both integrals and this process
is unfortunately standard but should not cause confusion,
since the angle bracket process always carries a subscript for time.

The form of~(\ref{eqn:eta_martingale}) and~(\ref{eqn:prelimit_martingale_variation})
is explained in Section~\ref{sec:heuristics}.
Note that since (juvenile) individuals are produced at rate $N \gamma \eta$, but each
has mass $1/N$,
these factors of $N$ cancel in~\eqref{eqn:eta_martingale}. 
Under our scaling, $N$ and $\theta=\theta(N)$ will tend to infinity in such a 
way that $\alpha:=\lim_{N\to\infty}\theta(N)/N$ exists and is finite.
From the expression~(\ref{eqn:prelimit_martingale_variation}) it is easy to guess that 
whether the limiting processes will be deterministic or stochastic
is determined by whether $\alpha$ is zero or nonzero.

@@

Since we take bounded $f$,
for most situations the bound $C_f(1+|f(x)|)$ in Condition~\ref{gamma_B_condition} above
can be safely replaced simply by $C_f$;
however, this will be useful
in certain situations where we consider a sequence of $f$ with increasing upper bounds.
We now give two concrete situations in which Condition~\ref{gamma_B_condition} is satisfied.
The proof is in Section~\ref{sec:preliminary_proofs}.

\begin{lemma}
    \label{lem:conditions on r}
    Assume that Conditions~\ref{def:model_setup} are satisfied,
    except for Condition~\ref{gamma_B_condition}.
    If either
\begin{enumerate}
\item
\label{control through r} 
    $|\partial_{x_i}r(x,\eta)|$ and 
    $|\partial_{x_ix_j}r(x,\eta)|$
    are uniformly bounded for $x\in\IR^d$, $\eta\in\measures$;
\item 
\label{control through gamma}
    or, $m^2\gamma(x,m)$ is uniformly bounded and there exists $C<\infty$
    such that for $\theta$ sufficiently large, and all $x\in\IR^d$, $\eta\in\measures$,
    \[
    \theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)q_\theta(x,dy)\leq C\smooth{\gamma}\eta(x),
    \]
    and
    \[\theta\int\big(\smooth{r}\eta(y)-\smooth{r}\eta(x)\big)^2q_\theta(x,dy)
    \leq C(\smooth{\gamma}\eta(x))^2 ,
    \]
\end{enumerate} 
    then Condition~\ref{gamma_B_condition} is also satisfied.
\end{lemma}

@@

\begin{theorem} \label{thm:nonlocal_convergence}
    Let $(\eta^N_t)_{t \geq 0}$
    be as defined in Definition~\ref{defn:mgale_construction}
	and assume that as $N \to \infty$, $\theta(N) \to \infty$
	in such a way that $\theta(N)/N \to \alpha$.
    (However, the kernels $\rho_r$, $\rho_\gamma$, and $\rho_F$
    remain fixed.) 
    Suppose that Assumptions~\ref{def:model_setup} hold and, 
    further, that $\{\eta_0^N\}_{N\geq 1}$ is a sequence of 
    purely atomic measures, with $\eta_0^N$ comprised of atoms of mass $1/N$,
    which is tight in $\measures$.
    Also assume there exists a nonnegative $f_0 \in C(\IR^d)$
    with uniformly bounded first and second derivatives
    (i.e., with $\sup_x \sup_{\|z\|=1} \sum_i \partial_{x_i} f_0(x) z_i$
    and $\sup_x \sup_{\|z\|=1} \sum_{ij} \partial_{x_i x_j} f_0(x) z_i z_j$ both finite)
    and $f_0(x) \to \infty$ as $|x| \to \infty$ for which
    $\langle f_0(x), \eta_0^N(dx)\rangle<C<\infty$
    for some $C$ independent of $N$.
    Then the sequence of processes
    $(\eta^N_t)_{t \geq 0}$ is tight, and for any limit point
    $(\eta_t)_{t \geq 0}$, for every $f \in C^\infty_b(\IR^d)$,
    \begin{align} \label{eqn:limiting_mgale_problem}
        \begin{split}
        M_t(f)
            &:=
            \langle f(x), \eta_t(dx) \rangle
            -
            \langle f(x), \eta_0(dx) \rangle
            \\ & \qquad
            -
            \int_0^t \big\langle
                \gamma(x, \eta_s)
                \mathcal{B}\left(
                    f(\cdot) r(\cdot, \eta_s)
                \right)(x)
                %\\ &\qquad \qquad \qquad {}
                +
                f(x)
                F(x, \eta_s),
                \eta_{s}(dx)
            \big\rangle ds
        \end{split}
    \end{align}
    is a martingale (with respect to the natural filtration),
    with angle bracket process
    \begin{align} \label{eqn:limiting_mgale_variation}
        \left\langle M(f) \right\rangle_t
        =
        \alpha
        \int_0^t
        \big\langle
            %\left(
	    2\gamma\left( x, \eta_{s} \right)
            r\left(x, \eta_{s} \right)% + \mu\left(x, \eta_{s} \right)\right)
            f^2(x),
            \eta_{s} (dx)
        \big\rangle ds. 
    \end{align}
    If $\alpha = 0$ the limit is deterministic.
\end{theorem}

@@

We now look at the second order term in~(\ref{generator prelimit}),
which will converge to the quadratic variation of the limiting process.
An individual at location $x$ gives birth 
to a surviving offspring at $y$ at rate
$$
\gamma(x,\eta) r(y,\eta) q_{\theta}(x, dy) ,
$$
and since this increments $\langle f, \eta \rangle$ by $f(y) / N$,
the contribution to the quadratic variation from birth events,
which occur at rate $\theta$ per individual 
(so, rate $N\theta |\eta|$ overall), is
$$
\int
    N \theta \gamma(x,\eta)
    \int \frac{1}{N^2} f^2(y) r(y,\eta)
    q_\theta(x,dy) 
\eta(dx) .
$$
Similarly, the increment in $\langle f, \eta\rangle$ resulting from 
the death of an individual at $x$ is $-f(x)/N$, and so combining with the 
above, the second order term in the generator takes the form
\begin{align*}
& G''(\langle f,\eta\rangle)
\frac{1}{2} N \theta
\left\{
    \int
        \gamma(x,\eta)
        \int \frac{1}{N^2}f^2(y)r(y,\eta)q_\theta(x,dy) 
    \eta(dx)
    +
    \int
        \mu_\theta(x,\eta)\frac{1}{N^2}f^2(x) 
    \eta(dx)
\right\} \\
&\qquad
= \frac{1}{2} G''(\langle f, \eta \rangle)
    \frac{\theta}{N}
	\int\left\{
		\gamma(x, \eta) \int f^2(y) r(y, \eta) q_\theta(x, dy) + f^2(x) \mu_\theta(x, \eta)
    \right\} \eta(dx) .
\end{align*}
Since $\int f^2(y) r(y, \eta) q_\theta(x, dy) \to f^2(x) r(x, \eta)$
and $r\gamma  + \mu_\theta = 2 r \gamma  - F / \theta \to 2 r \gamma$
as $\theta \to \infty$,
this converges to
\begin{align*}
\frac{\alpha}{2} G''(\langle f, \eta \rangle)
    \big\langle
	2 r(x, \eta) \gamma(x, \eta),
        \eta(dx)
    \big\rangle .
\end{align*}

@@

\begin{proof}
    First note that by Lemma~\ref{lem:eta_f_bound},
    $\IE[\langle 1, \eta^N_t \rangle] \le \IE[\langle 1, \eta^N_0 \rangle] e^{Ct}$
    for some $C$ (independent of $N$).
    Now, let $M^{N*}_t(f) = \sup_{0 \le s \le t} M^N_t(f)$,
    and as before let $\langle M^N(f) \rangle_t$ be the angle bracket process of $M^N_t(f)$.
    The Burkholder-Davis-Gundy inequality says that there is a $K$ for which
    $\IE\left[ M^{N*}_t(1) \right] \le K \IE[\sqrt{[M^N(1)]_t}]$,
    where $[M^N(1)]_t$ is the quadratic variation of $M^N(1)$.
    Furthermore, as discussed by \citet{hernandez/jacka:2022},
    % for $q \le 1$ there exist universal constants $c_q$ such that
    % for any local martingale with $M_0=0$, $\IE[[M]^{q}_T] \le c_q \IE[\langle M \rangle^q_T]$
    the expectation of the quadratic variation of a local martingale
    is bounded by a (universal) constant multiple of the expectation of its angle bracket process
    \citep[][Item (4.b'), Table 4.1, p. 162]{barlow/jacka/yor:1986}.
    Now, since $\sqrt{x} \le 1 + x$,
    in the notation of Lemma~\ref{lem:eta_f_bound}, there is a $C'$ such that
    \begin{align*}
        \IE\left[ M^{N*}_t(1) \right]
        &\le
        C'\left( 1 + \IE\left[ \left\langle M^N(1) \right\rangle_t \right] \right)
        \\ &=
        C'\left( 1 + \frac{\theta}{N} \IE\left[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\left\{
                \gamma(x, \eta^N_s)
                \int_{\IR^d} r(y, \eta^N_s) q_\theta(x, dy)
                + \mu_\theta(x, \eta^N_s)
            \right\}, \eta_s^N(dx)\Big\rangle
            ds
            \right] \right)
        \\ &=
        C'\Big( 1 + \IE\Big[
            \int_0^t
            %\int_{\IR^d} 
	    \Big\langle\Big\{
                \frac{2\theta}{N} \gamma(x, \eta^N_s) r(x, \eta^N_s)
        \\ &\qquad\qquad\qquad\qquad \qquad {}
		+ \frac{\gamma(x,\eta_s^N)}{N} 
		B^\theta_1(x, \eta^N_s)-\frac{1}{N}F(x,\eta^N_s)
            \Big\}, \eta_s^N(dx) \Big\rangle
            ds
            \Big] \Big) .
    \end{align*}

@@

To see why $\psi_t^{\epsilon,x}$ is useful,
first note that for any time-dependent function $\phi_t(x)$
with time derivative $\dot \phi_t(x) = \partial_t \phi_t(x)$,
\begin{multline}
\label{time dep mg prob}
    \langle \phi_t(x), \eta_t(dx) \rangle
    = 
    \langle \phi_0(x), \eta_0(dx) \rangle
    +
    M_t(\phi)
    +
    \int_0^t \big\langle \mathcal{L}^\theta \phi_s(x) 
        + \dot \phi_s(x), \eta_s(dx) \big\rangle ds
    \\  {}
    +
    \int_{0}^t \big\langle \phi_s(x) F(x, \eta_s) , \eta_s(dx) \big\rangle ds ,
\end{multline}
where $M_t(\phi)$ is a martingale (with respect to the natural filtration)
with angle bracket process given by~\eqref{eqn:prelimit_martingale_variation}
with $f$ replaced by $\phi_s(\cdot)$.
So, taking $\phi_s(\cdot)=\psi_{t-s}^{\epsilon,x}(\cdot)$ for $0 \le s \le t$,
\begin{eqnarray}
    \nonumber
    \rho_\epsilon*\eta_t(x)&=&\langle\psi^{\epsilon,x}_0(y),\eta_t(dy)\rangle
    \\
    &=&\langle\psi^{\epsilon,x}_t(y),\eta_0(dy)\rangle
    +\int_0^t\big\langle\psi^{\epsilon,x}_{t-s}(y)
    F\big(\rho_\epsilon*\eta_s(y)\big), \eta_s(dy)\big\rangle ds
    +M_t(x),
    \label{expn rhoepsilon} 
\end{eqnarray}
where $M_t(x)$ has mean zero and a second moment we can easily write down.

@@

We follow the approach of Lemma~\ref{lem:eta_compact_containment}.
Consulting~\eqref{eqn:prelimit_martingale_variation},
the angle bracket process of $\widehat{M}^N(f)$ satisfies
$ \IE[\langle\widehat{M}^N_f\rangle_T] \le C (\theta/N) \int_0^T \IE[\langle 1, \eta_s \rangle] ds \le C' \theta/N$
for some constants $C$ and $C'$.
Now, using the Burkholder-Davis-Gundy inequality and \citet{hernandez/jacka:2022},
$\IE[\sup_{0 \le u \le T} |\widehat{M}^N(f)_u|^2] \le C'' \IE[\langle\widehat{M}^N(f)\rangle_u]$,
and so using Markov's inequality,
\begin{equation}
\label{martingale term to zero}
    \limsup_{N\to\infty}\IP\Big[
        2\sup_{0\leq u\leq T}|\widehat{M}^N(f)_u|
        >\frac{\nu}{3}
    \Big]
    \leq
    \limsup_{N\to\infty} \frac{36}{\nu^2} C'' \IE\big[\langle\widehat{M}^N(f)\rangle_T\big]
    \leq
    \limsup_{N\to\infty}\frac{36}{\nu^2}\frac{C' C''\theta}{N\epsilon^d} 
    =0. 
\end{equation}
